%-------------------------
% Resume in Latex
% Author : Jake Gutierrez
% Based off of: https://github.com/sb2nov/resume
% License : MIT
%------------------------

\documentclass[letterpaper,10.99999999999999pt]{article}

\usepackage{latexsym}
\usepackage[empty]{fullpage}
\usepackage{titlesec}
\usepackage{marvosym}
\usepackage[usenames,dvipsnames]{color}
\usepackage{verbatim}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyhdr}
\usepackage[english]{babel}
\usepackage{tabularx}
\input{glyphtounicode}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=blue,
}

\urlstyle{same}

%----------FONT OPTIONS----------
% sans-serif
% \usepackage[sfdefault]{FiraSans}
% \usepackage[sfdefault]{roboto}
% \usepackage[sfdefault]{noto-sans}
% \usepackage[default]{sourcesanspro}

% serif
% \usepackage{CormorantGaramond}
% \usepackage{charter}


\pagestyle{fancy}
\fancyhf{} % clear all header and footer fields
\fancyfoot{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}

% Adjust margins
\addtolength{\oddsidemargin}{-0.5in}
\addtolength{\evensidemargin}{-0.5in}
\addtolength{\textwidth}{1in}
\addtolength{\topmargin}{-.5in}
\addtolength{\textheight}{1.0in}

\raggedbottom
\raggedright
\setlength{\tabcolsep}{0in}

% Sections formatting
\titleformat{\section}{
  \vspace{-8pt}\scshape\raggedright\large
}{}{0em}{}[\color{black}\titlerule \vspace{-5pt}]

% Ensure that generate pdf is machine readable/ATS parsable
\pdfgentounicode=1

%-------------------------
% Custom commands
\newcommand{\resumeItem}[1]{
  \item\small{
    {#1 \vspace{-1pt}}
  }
}

\newcommand{\resumeSubheading}[4]{
  \vspace{-1pt}\item
    \begin{tabular*}{0.97\textwidth}[t]{l@{\extracolsep{\fill}}r}
      \textbf{#1} & #2 \\
      \textit{\small#3} & \textit{\small #4} \\
    \end{tabular*}\vspace{-6pt}
}

\newcommand{\resumeSubSubheading}[2]{
    \item
    \begin{tabular*}{0.97\textwidth}{l@{\extracolsep{\fill}}r}
      \textit{\small#1} & \textit{\small #2} \\
    \end{tabular*}\vspace{-6pt}
}

\newcommand{\resumeProjectHeading}[2]{
    \item
    \begin{tabular*}{0.97\textwidth}{l@{\extracolsep{\fill}}r}
      \small#1 & #2 \\
    \end{tabular*}\vspace{-6pt}
}

\newcommand{\resumeEducationEntry}[3]{
  \vspace{-2pt}\item
    \begin{tabular*}{0.97\textwidth}[t]{l@{\extracolsep{\fill}}r}
      \textbf{#1} & \\
      \textit{\small#2} & \textit{\small #3} \hfill
    \end{tabular*}\vspace{-4pt}
}

\newcommand{\resumeSingleLineSubheading}[2]{
  \item
    \begin{tabular*}{0.97\textwidth}[t]{l@{\extracolsep{\fill}}r}
      \textbf{#1} & \textit{\small#2} \\
    \end{tabular*}\vspace{-6pt}
}



\newcommand{\resumeSubItem}[1]{\resumeItem{#1}\vspace{-4pt}}

\renewcommand\labelitemii{$\vcenter{\hbox{\tiny$\bullet$}}$}

\newcommand{\resumeSubHeadingListStart}{\begin{itemize}[leftmargin=0.15in, label={}]}
\newcommand{\resumeSubHeadingListEnd}{\end{itemize}}
\newcommand{\resumeItemListStart}{\begin{itemize}}
\newcommand{\resumeItemListEnd}{\end{itemize}\vspace{-5pt}}

%-------------------------------------------
%%%%%%  RESUME STARTS HERE  %%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{document}

%----------HEADING----------
% \begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%   \textbf{\href{http://sourabhbajaj.com/}{\Large Sourabh Bajaj}} & Email : \href{mailto:sourabh@sourabhbajaj.com}{sourabh@sourabhbajaj.com}\\
%   \href{http://sourabhbajaj.com/}{http://www.sourabhbajaj.com} & Mobile : +1-123-456-7890 \\
% \end{tabular*}

\begin{center}
    \textbf{\Huge \scshape \underline{Aidan Andrews}} \\ \vspace{1pt}
    \small +1 (310) 910-4721 $|$ \href{mailto:aidansa2@illinois.edu}{aidansa2@illinois.edu} $|$
    \href{https://www.linkedin.com/in/aidanandrewss/}{linkedin.com/in/aidanandrewss/} $|$
    \href{https://github.com/aidanandrews22}{github.com/aidanandrews22} $|$
    \href{https://aidanandrews.info}{aidanandrews.info}
\end{center}

%-----------EDUCATION-----------

\section{Education}
  \resumeSubHeadingListStart
    \resumeEducationEntry
      {University of Illinois at Urbana-Champaign}
      {Bachelor of Science in Engineering Physics: AI/ML and Quantum Computing Track}
      {Expected Graduation: 2026, GPA: 3.9}
  \resumeSubHeadingListEnd

%-----------EXPERIENCE-----------
\section{Experience}
  \resumeSubHeadingListStart

  \resumeSubheading
      {AI/ML Research Intern}{Nov 2024 – Present}
      {AIFARMS National AI Institute \& Center for Digital Agriculture}{On-Site}
      \resumeItemListStart
        \resumeItem{Developing tools for LLM integration in production-scale chatbots, enhancing agentic workflows, which facilitated over 10,000 conversations and the efficient resolution of over 13,500 real-world problems.}
        \resumeItem{Implementing AI-driven solutions to increase the efficiency of current agricultural practices at scale.}
        \resumeItem{Building and researching agents, RFT, RAG, Reasoning, and pre/post training to advance AI capabilities in agriculture.}
      \resumeItemListEnd

  \resumeSubheading
      {Startup (Signed NDA)}{May 2024 – Aug 2024}
      {Founding Machine Learning Engineer}{Remote}
      \resumeItemListStart
        \resumeItem{Engineered advanced NLP systems, optimizing Retrieval-Augmented Generation (RAG) and intent classification.}
        \resumeItem{Designed a novel "wavular" RAG approach and hybrid embedding-based classification system.}
        \resumeItem{Applied vector space models, similarity metrics (cosine, Euclidean), and text representation methods.}
        \resumeItem{Managed large-scale NLP datasets with complex preprocessing, delivering innovative solutions for NLP challenges in resource-constrained environments.}
      \resumeItemListEnd
      
    \resumeSubheading
      {\href{https://www.aaxiom.org/}{AAXIOM}}{March 2023 – Present}
      {FOUNDER, Sole Developer}{}
      \resumeItemListStart
        \resumeItem{Founded a parent company to oversee all startups and projects I develop. NVRALONE achieved \$10K/month in profit, IlliniPlan reached 1K beta users, and IlliniSpots is projected to reach 12K users based on survey data.}
      \resumeItemListEnd
      
% -----------Multiple Positions Heading-----------
%    \resumeSubSubheading
%     {Software Engineer I}{Oct 2014 - Sep 2016}
%     \resumeItemListStart
%        \resumeItem{Apache Beam}
%          {Apache Beam is a unified model for defining both batch and streaming data-parallel processing pipelines}
%     \resumeItemListEnd
%    \resumeSubHeadingListEnd
%-------------------------------------------

    % \resumeSubheading
    %   {Amara Andrews Mayoral Campaign}{May. 2022 -- Nov. 2022}
    %   {Mayoral Campaign Volunteer}{Cedar Rapids, IA}
      % \resumeItemListStart
      %   \resumeItem{Provided strategic recommendations for community outreach and voter engagement.}
      %   \resumeItem{Contributed to campaign planning and execution, including social media strategies and event planning.}
    % \resumeItemListEnd

    % \resumeSubheading
    %   {Project Manager for CS124 Honors and Course Assistant CS124}{Dec. 2023 -- May. 2024}
    %   {}{}
    % \resumeSubheading
    %   {}{Dec. 2023 -- May. 2024}
    %   {}{The University of Illinois at Urbana-Champaign}

% End of the subheading list
    \resumeSubHeadingListEnd

%-----------AI/ML PROJECTS-----------
\section{Projects}
    \resumeSubHeadingListStart
    
        \resumeProjectHeading
          {\textbf{\href{https://www.aidanandrews.info/projects/climatepredict}{Climate Predict: Global Warming Calculator}} $|$ \emph{Rust, Regression Models}}{}
          \resumeItemListStart
            \resumeItem{Developed a multi regression model in Rust. The project comprises 3 regression models: linear regression, polynomial regression, and random forest regression (incomplete).}
            \resumeItem{Used public NASA datasets to predict temperature change given CO2 emmisions.}
          \resumeItemListEnd
          
      \resumeProjectHeading
          {\textbf{\href{https://www.aidanandrews.info/projects/devanagari-web-app}{Devanagari Language Learning App}} $|$ \emph{Python, Flask, TensorFlow, Keras, scikit-learn, Pandas, NumPy, Matplotlib}}{}
          \resumeItemListStart
            \resumeItem{Engineered a Convolutional Neural Network (CNN) to accurately recognize natural handwritten language.}
            \resumeItem{Enhanced model performance and accuracy through data preprocessing and augmentation techniques, ensuring robustness against diverse handwriting styles.}
            \resumeItem{Optimized training efficiency with back-propagation and gradient descent.}
          \resumeItemListEnd

      \resumeProjectHeading
        {\textbf{\href{https://www.aidanandrews.info/projects/voxed}{Voxed: AI Powered Notebook}} $|$ \emph{Google Cloud Services, OpenAI API, Ubuntu, PyDub, Python}}{}
        \resumeItemListStart
          \resumeItem{Developed full-stack AI app that allows users to upload any file type (including video) and create a notebook where they can interact with their files via llm.}
        \resumeItemListEnd
        
        \resumeProjectHeading
        {\textbf{\href{https://www.aidanandrews.info/projects/illini-plan}{IlliniPlan: Student Course Scheduler and Degree Progress}} $|$ \emph{Postgres, React, Typescript, Tailwind}}{}
        \resumeItemListStart
        \resumeItem{Developed algorithms for degree progress tracking, sequence validation, course recommendations, and search. Optimized database calls for performance.}
        \resumeItemListEnd
        
        \resumeProjectHeading
        {\textbf{\href{https://www.aidanandrews.info/projects/illini-spots}{IlliniSpots: UIUC Study Spot Finder (IOS App)}} $|$ \emph{Postgres, Firebase, Swift, React Native, Expo, Search}}{}
        
        \resumeProjectHeading
        {\textbf{\href{https://github.com/aidanandrews22/note-taking-app}{Productivity App (notes, todos, calendar)}} $|$ \emph{React, Swift, AWS (lambda, amplify, gateway), Firebase, Typescript, JS, tailwind}}{}
        \resumeProjectHeading
        {\textbf{\href{https://github.com/aidanandrews22/Portfolio}{Portfolio Website}} $|$ \emph{React, AWS, Github API, Typescript, JS, tailwind}}{}
        \resumeProjectHeading
            {\textbf{\href{https://github.com/aidanandrews22/VoxAI-physical}{VoxAI: The Autonomous Learning Assistant}} $|$ \emph{Google Cloud Services, OpenAI API, Ubuntu, PyDub, Python}}{}
        
    \resumeSubHeadingListEnd

%-----------EXTRACURRICULAR ACTIVITIES-----------
\section{Extracurricular Activities}
  \resumeSubHeadingListStart
        \resumeProjectHeading
          {\textbf{Division 1 College Ice Hockey} $|$ \emph{University Of Illinois Men's ACHA Division 1 Hockey}}{}
        \resumeProjectHeading
          {\textbf{Developed App for Plastic Surgeon} $|$ \emph{Facilitates patient database management and surgical procedure simulations}}{}
        \resumeProjectHeading
          {\textbf{Chess} $|$ \emph{Rated 2000 over-the-board, Training to become a FIDE Master}}{}
        \resumeProjectHeading
          {\textbf{BuildIllinois Cohort Team} $|$ \emph{Building and fostering startups at UIUC}}{}
  \resumeSubHeadingListEnd

%
%-----------SKILLS & OTHER-----------
\begin{minipage}[t]{0.48\textwidth}
\section{Technical Skills}
 \begin{itemize}[leftmargin=0.15in, label={}]
    \small{\item{
     \textbf{Languages in order of proficiency}{: C++, Python, \newline TypeScript, Kotlin, JavaScript, Java} \\
     \textbf{Developer Tools}{: Git, Linux, CLI, Docker, AWS,} \\
     \hspace*{0em}{Pytorch, Tensorflow, Firebase, Transformers, HuggingFace, Supabase, Postgres} \\
    }}
 \end{itemize}
\end{minipage}
\hfill
\vrule
\hfill
\begin{minipage}[t]{0.48\textwidth}
\section{Other}
 \begin{itemize}[leftmargin=0.15in, label={}]
    \small{\item{
    \textbf{Clubs/Organizations}{: James Scholar, NSBE, MEP} \\
    \textbf{Interesting Facts}{: Streamed myself working 80+ hours a week on youtube with 11k+ followers, 1M+ views. \newline Own and run an educational and motivational discord server with over 1k members focused on CS, ML/AI, Math, Physics, education, and overall life.} \\
    }}
 \end{itemize}
\end{minipage}
%-------------------------------------------
\end{document}
