\documentclass{article}
\usepackage{amsmath, amssymb, bm}
\usepackage{braket}

\begin{document}

\section{Matrix Representation of Quantum Gates}
In quantum computing, the state of an \(n\)-qubit system is represented by a \(2^n\)-dimensional vector. Quantum gates are represented by \(2^n \times 2^n\) unitary matrices. A six-qubit quantum gate must be represented by a \(64 \times 64\) unitary matrix. The unitary condition:
\[
U^{\dagger}U = U U^{\dagger} = I,
\]
where \(U^{\dagger}\) is the Hermitian conjugate (conjugate transpose) of \(U\).

\section{Pauli Matrices as Quantum Gates}
The Pauli matrices (\(X, Y, Z\)) are fundamental single-qubit gates:
\begin{align*}
X & = \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix} \quad \text{(Bit-flip gate)} \\
Y & = \begin{bmatrix} 0 & -i \\ i & 0 \end{bmatrix} \quad \text{(Bit \& Phase-flip)} \\
Z & = \begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix} \quad \text{(Phase-flip gate)}
\end{align*}`'
These matrices are unitary (\(U^{\dagger} = U^{-1}\)) and Hermitian (\(U = U^{\dagger}\)), ensuring valid quantum operations.

\section{Controlled-NOT (CNOT) Gate}
The CNOT gate acts on two qubits, with one acting as a control and the other as a target. It performs a NOT operation on the target if and only if the control qubit is \(|1\rangle\). The matrix representation:
\[
\text{CNOT} = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 & 1 \\ 0 & 0 & 1 & 0 \end{bmatrix}
\]
Using Dirac notation, its action is:
\begin{align*}
|00\rangle &\to |00\rangle \\
|01\rangle &\to |01\rangle \\
|10\rangle &\to |11\rangle \quad \text{(flips target qubit)} \\
|11\rangle &\to |10\rangle
\end{align*}
The CNOT gate can be used to create Bell states when combined with the Hadamard gate.

\section{Hadamard Gate and Bell State Generation}
The Hadamard gate (\(H\)) puts a qubit into an equal superposition:
\[
H = \frac{1}{\sqrt{2}} \begin{bmatrix} 1 & 1 \\ 1 & -1 \end{bmatrix}
\]
Used to create Bell states in combination with CNOT. Example Bell state:
\[
|\beta_{00}\rangle = \frac{|00\rangle + |11\rangle}{\sqrt{2}}
\]
The circuit consists of:
\begin{itemize}
    \item Applying \(H\) to the first qubit.
    \item Applying CNOT, with the first qubit as the control.
\end{itemize}

\section{Quantum Circuit Representation}
\begin{itemize}
    \item Wires represent qubits and their evolution through time, not physical movement.
    \item Single-qubit gates (\(X, H, Z\), etc.) act on one qubit.
    \item Multi-qubit gates (CNOT, Toffoli) involve entanglement.
    \item Measurement collapses the quantum state and is usually represented by an encircled \(M\) in diagrams.
\end{itemize}

\section{Controlled Gates and Universal Quantum Gates}
Controlled gates allow conditional operations:
\begin{itemize}
    \item Controlled-Z (CZ): Applies a \(Z\) gate to the target when control is \(|1\rangle\).
    \item Controlled-Hadamard (CH): Applies \(H\) to the target when control is \(|1\rangle\).
    \item Toffoli Gate (CCNOT): Uses two control qubits and one target, flipping the target only when both controls are \(|1\rangle\).
    \item Fredkin Gate (CSWAP): Swaps two qubits based on a control qubit.
\end{itemize}

\section{Gate Decomposition}
Complex multi-qubit gates can be decomposed into single-qubit gates and CNOTs. Example: Controlled-\(U\) decomposition replaces a controlled-\(U\) gate with a sequence of CNOTs and single-qubit rotations, used for efficient quantum circuit optimization.

\section{No-Cloning Theorem and Quantum Cloning}
The CNOT gate cannot clone an arbitrary quantum state due to the no-cloning theorem. Example: Trying to copy \(\alpha|0\rangle + \beta|1\rangle\) results in entanglement, not an identical copy. Some specific states, like \(|0\rangle\) or \(|1\rangle\), can be cloned, but arbitrary superpositions cannot.

\section{Key Takeaways for Class Questions}
\begin{itemize}
    \item What size matrix represents an \(n\)-qubit gate? \(2^n \times 2^n\) unitary matrix.
    \item Why are Pauli matrices quantum gates? They are unitary and Hermitian.
    \item What is the role of CNOT? Flips the target qubit if the control is \(|1\rangle\).
    \item How does Hadamard create superposition? Equal probability of \(|0\rangle\) and \(|1\rangle\).
    \item What do wires in a circuit mean? Time evolution of a quantum state.
    \item Can quantum states be cloned? No, due to the no-cloning theorem.
\end{itemize}

\end{document}
